---
title: "Week 4"
author: "Dr. Harsh Pradhan"
affiliation: "Institute of Management Studies, Banaras Hindu University"
format: 
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    theme: cosmo
    css: styles.css
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    fontsize: 11pt
    documentclass: article
bibliography: references.bib
---

## Introduction

This eBook is a comprehensive companion to the course *Basic Statistics using GUI-R (RKWard)*. It includes foundational theory, practical examples, and step-by-step explanations, with integrated GUI-R usage.

## Course Overview

### Course Name

**Basic Statistics using GUI-R (RKWard)**

### Instructor Profile

Dr. Harsh Pradhan is Assistant Professor at the Institute of Management Studies, Banaras Hindu University.  
📎 [Faculty Profile](https://bhu.ac.in/Site/FacultyProfile/1_5?FA000562)

### Learning Objectives

- Understand core concepts in statistics
- Apply t-tests and ANOVA using real data
- Compute confidence intervals and test statistics
- Use GUI-R (RKWard) for statistical analysis

## Chapter 1: Fundamental Concepts

### Descriptive Statistics

#### Central Tendency

- **Mean**  
- **Median**  
- **Mode**  

#### Dispersion

- **Range**  
- **Variance**  
- **Standard Deviation**

#### Example:

```{r}
data <- c(4, 8, 6, 5, 3)
mean(data)
median(data)
sd(data)
```

### Standard Error

$$
SE = \frac{s}{\sqrt{n}}
$$

Small SE = sample mean is a good estimate of the population mean.

### Central Limit Theorem

For $n > 30$, sampling distribution of the mean approximates normal:

$$
\bar{X} \sim \mathcal{N}(\mu, \frac{\sigma}{\sqrt{n}})
$$

### Confidence Intervals

$$
CI = \bar{x} \pm Z \cdot \frac{s}{\sqrt{n}}
$$

Interpret 95% CI as: 95 of 100 such intervals would contain the true mean.

## Chapter 2: Estimation

### Types of Estimates

| Type            | Description          | Example        |
|-----------------|----------------------|----------------|
| Point Estimate  | Single value         | Sample mean    |
| Interval Estimate| Range + confidence | Confidence Int |

### Parameter vs Statistic

| Term       | Description                            |
|------------|----------------------------------------|
| Parameter  | Value from population (e.g., $\mu$)    |
| Statistic  | Value from sample (e.g., $\bar{x}$)    |

## Chapter 3: Hypothesis Testing

- **Null Hypothesis ($H_0$)**: No effect  
- **Alternative Hypothesis ($H_1$)**: Some effect  
- **Type I Error**: Reject $H_0$ when true  
- **Type II Error**: Fail to reject $H_0$ when false

## Chapter 4: Student's T-Test

### Types

| Test Type         | Description                    |
|-------------------|--------------------------------|
| One-Sample        | Compare sample to fixed value  |
| Independent       | Compare two unrelated groups   |
| Paired            | Compare two related groups     |

### One-Sample T-Test Example

```{r}
data <- c(22, 24, 27, 26, 28, 23, 25, 29, 21, 26, 24, 27)
t.test(data, mu = 25)
```

### Test Statistic

$$
t = \frac{\bar{x} - \mu}{SE}
$$

### Degrees of Freedom

$$
df = n - 1
$$

### Decision Rule

Compare calculated $t$ to table value. If $|t| > t_{critical}$, reject $H_0$.

### T-Test in GUI-R

1. Import data  
2. Choose T-Test  
3. Define groups  
4. Run & interpret output

## Chapter 5: ANOVA

### Purpose

Used when comparing means across 3+ groups.

#### One-Way ANOVA Formula

$$
F = \frac{MS_{between}}{MS_{within}}
$$

Where:

- $MS_{between} = \frac{SS_{between}}{df_{between}}$  
- $MS_{within} = \frac{SS_{within}}{df_{within}}$  

#### Assumptions

- Normality  
- Homogeneity of variance  
- Independence  

#### Example Table

| Group | Mean | Var | n  |
|-------|------|-----|----|
| A     | 5.5  | 1.5 | 30 |
| B     | 7.1  | 2.0 | 30 |
| C     | 6.8  | 1.8 | 30 |

### Post-Hoc Tests

Run if ANOVA is significant to locate pairwise differences.

### ANOVA in GUI-R

1. Load data  
2. Choose "One-Way ANOVA"  
3. Define groups  
4. Interpret output

## Chapter 6: GUI-R Workflow

1. **Import Data** (CSV, Excel)  
2. **Choose Test** (T-Test, ANOVA, etc.)  
3. **Run** the analysis  
4. **Interpret** the output  
5. **Export** the results or visualizations

## Chapter 7: Advanced Concepts

### Variance Partitioning

$$
\text{Total Variance} = \text{Explained Variance} + \text{Unexplained Variance}
$$

| Explained Terms     | Unexplained Terms   |
|---------------------|---------------------|
| Systematic          | Random              |
| Predictive          | Error               |
| Deterministic       | Noise               |

### Degrees of Freedom

For equation $x + y + z = 3$, if 2 values are known, third is fixed.  
Hence, $df = n - k$ where $n$ = total variables, $k$ = constraints.

### Chi-Square and F Distribution

- **Chi-Square**: Categorical variable comparison  
- **F-Distribution**: Used in ANOVA, variance testing

### Univariate, Bivariate, Multivariate

| Type        | Variables | Example                         |
|-------------|-----------|---------------------------------|
| Univariate  | 1         | Height                          |
| Bivariate   | 2         | Height vs Weight                |
| Multivariate| >2        | Study w/ Age, Gender, Income    |

### Parametric Test Assumptions

- Interval/Ratio DV  
- Random Sampling  
- Normality  
- Equal Variances  

If assumptions violated → use non-parametric test.

### Effect Size

$$
\text{Effect Size} = \frac{|\mu_1 - \mu_2|}{\sigma}
$$

Used for comparison across studies.

### Power of a Test

$$
\text{Power} = 1 - \beta
$$

Higher power → lower chance of Type II error  
Power increases with sample size, effect size

## Conclusion

Statistics is the language of data. GUI-R makes statistical tools accessible for everyone. This book empowers you to analyze data effectively using t-tests, ANOVA, and confidence intervals in a GUI environment.

## References

- Pradhan, H. (2023). *Basic Statistics using GUI-R (RKWard)*  
- Field, A. (2013). *Discovering Statistics Using R*.  
- [https://methods.sagepub.com](https://methods.sagepub.com)  

## Chapter 8: Advanced T-Test Applications

### Paired Sample T-Test

Used when the same group is measured twice (e.g., before and after).

#### Example:

```{r}
before <- c(80, 82, 79, 84, 88)
after <- c(78, 81, 76, 83, 86)
t.test(before, after, paired = TRUE)
```

### Independent Samples T-Test

Compare means of two unrelated groups.

```{r}
group1 <- c(85, 90, 88, 92, 87)
group2 <- c(80, 83, 85, 84, 82)
t.test(group1, group2)
```

### One-Sample T-Test with GUI-R

- Import dataset
- Use 'Descriptive Statistics' to check mean
- Navigate to 'T-Test' → 'One Sample'
- Input hypothesized mean and run

## Chapter 9: More on Confidence Intervals

### Visualizing Confidence Intervals in R

```{r}
x <- c(88, 90, 85, 87, 89)
mean_x <- mean(x)
se <- sd(x) / sqrt(length(x))
ci_lower <- mean_x - 1.96 * se
ci_upper <- mean_x + 1.96 * se
c(ci_lower, ci_upper)
```

Plot using `ggplot2`:

```{r}
library(ggplot2)
df <- data.frame(x = x)
ggplot(df, aes(y = x, x = 1)) +
  geom_point() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1)
```

## Chapter 10: Robust ANOVA Models

### Two-Way ANOVA

Examines the effect of two categorical independent variables on a continuous dependent variable.

```{r}
# Sample dataset for demonstration
dataset <- data.frame(
  score = c(85, 90, 88, 92, 87, 80, 83, 85, 84, 82),
  gender = rep(c("Male", "Female"), each = 5),
  teaching_method = rep(c("A", "B"), times = 5)
)
aov_result <- aov(score ~ gender * teaching_method, data = dataset)
summary(aov_result)
```

### Repeated Measures ANOVA

Use when the same subjects are used for each treatment.

```{r}
# Sample repeated measures data in long format
data_long <- data.frame(
  id = rep(1:5, each = 3),
  condition = rep(c("A", "B", "C"), times = 5),
  score = c(85, 88, 90, 80, 82, 85, 78, 80, 83, 90, 92, 95, 88, 90, 91)
)
library(ez)
ezANOVA(data = data_long, dv = .(score), wid = .(id), within = .(condition))
```

## Chapter 11: Effect Size Measures

### Cohen's d

$$
d = \frac{\bar{x}_1 - \bar{x}_2}{s_p}
$$

Where $s_p$ is the pooled standard deviation.

#### R Example

```{r}
library(effsize)
cohen.d(group1, group2)
```

### Eta-Squared ($\eta^2$)

Used for ANOVA:

$$
\eta^2 = \frac{SS_{between}}{SS_{total}}
$$

## Chapter 12: Statistical Assumptions Checking

### Normality

Use Shapiro-Wilk test:

```{r}
# Sample data frame for normality test
data <- data.frame(variable = c(88, 90, 85, 87, 89, 91, 92, 88, 90, 87))
shapiro.test(data$variable)
```

Visualize:

```{r}
qqnorm(data$variable)
qqline(data$variable)
```

### Homogeneity of Variance

Use Levene's Test:

```{r}
# Sample data frame for Levene's Test
data <- data.frame(
  variable = c(88, 90, 85, 87, 89, 91, 92, 88, 90, 87),
  group = rep(c("A", "B"), each = 5)
)
library(car)
leveneTest(variable ~ group, data = data)
```

## Chapter 13: Non-Parametric Alternatives

### Wilcoxon Signed Rank Test

```{r}
wilcox.test(before, after, paired = TRUE)
```

### Mann-Whitney U Test

```{r}
wilcox.test(group1, group2)
```

### Kruskal-Wallis Test

Non-parametric alternative to ANOVA.

```{r}
# Sample data frame for Kruskal-Wallis Test
data <- data.frame(
  score = c(85, 88, 90, 80, 82, 85, 78, 80, 83, 90, 92, 95, 88, 90, 91),
  group = rep(c("A", "B", "C"), times = 5)
)
kruskal.test(score ~ group, data = data)
```

## Visualizing Statistical Results

## Boxplots

```{r}
ggplot(data, aes(x = group, y = score)) +
  geom_boxplot()
```

### Histograms

```{r}
ggplot(data, aes(x = score)) +
  geom_histogram(binwidth = 2)
```

### Density Plot

```{r}
ggplot(data, aes(x = score)) +
  geom_density()
```

## RKWard (GUI-R) Tips

- Use menu-based analysis for beginners
- Save and export plots easily
- Integrate with R scripts for reproducibility

### 📘Summary: Basic Statistics using GUI-R (RKWard)

This eBook, authored by Dr. Harsh Pradhan (Assistant Professor at the Institute of Management Studies, Banaras Hindu University), serves as a comprehensive guide to understanding and applying basic statistical concepts, particularly in the GUI-based software RKWard (GUI-R).

Key Highlights:
1. Descriptive Statistics
Covers measures of central tendency (mean, median, mode) and variability (range, variance, standard deviation).
Introduces standard error and its role in estimating population parameters.
2. Inferential Statistics
Introduces the Central Limit Theorem and how it forms the foundation for many statistical techniques.
Confidence intervals are explained both theoretically and with practical calculations.
3. T-Tests (Student's t)
Explains one-sample, independent-sample, and paired-sample t-tests.
Includes step-by-step computation and GUI-R implementation.
Includes interpretation of p-values, degrees of freedom, and test statistics.
4. Analysis of Variance (ANOVA)
Covers one-way, two-way, and repeated measures ANOVA.
Focuses on the F-statistic, assumptions, and post-hoc analyses.
Discusses partitioning of variance into systematic and unsystematic components.
5. Effect Size and Statistical Power
Introduces Cohen's d, eta-squared, and power analysis.
Emphasizes that statistical significance does not always imply practical importance.
6. Assumption Testing
Tests for normality (Shapiro-Wilk, QQ plot).
Tests for homogeneity of variance (Levene's test).
Highlights when to use non-parametric alternatives.
7. Non-Parametric Tests
Introduces Wilcoxon signed-rank, Mann-Whitney U, and Kruskal-Wallis tests as robust alternatives to parametric methods.
8. Data Visualization in R
Demonstrates use of boxplots, histograms, and density plots using ggplot2.
Provides example R code for reproducibility.
9. GUI-R (RKWard) Usage
Offers practical steps for using GUI-R for all statistical techniques covered.
Designed to bridge the gap for learners unfamiliar with command-line R.

